Beyond clines: lineages and haplotype blocks in hybrid zones

### 2016-03-05 drafts/hybrid_zone.tex
notes on refs

\alisa{Better summary of what Barton did.
Population genetic modeling of hybrid zones has focused largely on the 
equilibrium states of clines at selected loci \citep{Nagylaki1975,Barton1979a,Barton1986,Slatkin1973,Slatkin1975,May1975,Haldane1948}.  
Since stronger selection reduces the amount of genetic mixing,
these models predict the narrowest clines 
(i.e., the most rapid changes in allele frequencies)
at loci close to selected loci, and wider clines at less tightly linked loci.
In particular \citet{Barton1979,Barton1986} have described patterns of clinal variation at loci linked to targets of selection.  
This provides some expectation for how selection shapes covariance in ancestry at a single locus over geographic space.
Our aim here is to extend this to include expectations for how selection shapes covariance in ancestry across the genome.}

@@

\paragraph{Establishment of the cline at the selected locus}
Deterministic theory predicts that after secondary contact,
the two alleles at the selected locus will form a stable cline \citep{Barton1979},
affecting neighboring loci as well.
First, we turn our attention to how the cline at the selected locus itself is formed.
Let $p(x,t)$ denote the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and that $p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
% (In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
Let $\sigma^2$ be the mean squared distance between parent and offspring 
(i.e., the variance of the distribution of dispersal distance).
Assuming that: (i) alleles locally assort into diploids randomly, 
(ii) habitat and dispersal are homogeneous, and 
(iii) $s$ and $\sigma$ are small, then the commonly-used equation that $p$ solves is
\begin{align} \label{eqn:cline_pde}
    \dot p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\dot p(t,x)$ is the time derivative and $\Delta p$ is the Laplacian,
as derived in \citet{Bazykin1969}; for discussion of diploidy see \citet{christiansen1995genotypic}.
Although there is no known exact analytic solution to this equation, even in steady state, 
a good approximation to the equilibrium solution is that
$\lim_{t \to \infty} p(x,t) \approx (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}. 
This relies on several approximations, but the general conclusions should be quite robust: 
the stable cline has width of order $\sigma/\sqrt{s}$, and decays exponentially. 
Note that rescaling space and time by $\sigma/s$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} results in a dimensionless equation,
implying that the cline establishes over a timescale of the order $1/s$. 
In other words, the cline is established through diffusion, despite being slowed down somewhat by selection against heterozygotes.

>>>>

%%%
%% these deal with environmental variation in fitness
% Haldane1948 : does 1D stable cline including dominance for step fitness gradient
% Fisher1950 : continuous gradient in selection
% Hansen1966 : does numerical solns to discretized ODE including observing minimal patch size
% Nagylaki1975 : writes down equations in 2D and conditions for stable solns
%%%
%% these deal with hybrid zones
% Bazykin1969 : writes down the stable solution for the one-dimensional cline
% barton1979dynamics of hybrid zones: looks at how hybrid zones move using a potential; also looks at the strength of stochastic effects. seems to NOT say how fast they relax?
% barton1979gene flow past a cline : step in frequency of linked neutral allele 
%    proportional to gradient in frequency and inverse to recomb rate
%    adaptive introgression occurs quickly 
% barton1986 and bengtsson : looks at many linked incompatibilities
%    with barrier B, neutral allele delayed by (B/sigma)^2, 
%    while adaptively introgressing allele delayed only if (B/sigma)^2 > 1/s.
%    also does size of step of allele frequency
% barton2008effect of a barrier to gene flow : on fluctuations in allele frequency, using diffusion approximation; 
%    finds is more sensitive in 2d than 1d
%%%
%% this deals with both:
% Slatkin1973 : does environmental cline and minimal width of a patch;
%    shows discontinuous freqs at barriers to dispersal
%    AND observes that het disadvantage leads to steeper clines,
%    so maybe environment determines location of cline; het disadvantage the shape
% May1975 : spatial resolution of adaptation; also does het disadvantage
\alisa{Better summary of what Barton did.
Population genetic modeling of hybrid zones has focused largely on the 
equilibrium states of clines at selected loci \citep{Nagylaki1975,Barton1979a,Barton1986,Slatkin1973,Slatkin1975,May1975,Haldane1948}.  
Since stronger selection reduces the amount of genetic mixing,
these models predict the narrowest clines 
(i.e., the most rapid changes in allele frequencies)
at loci close to selected loci, and wider clines at less tightly linked loci.
In particular \citet{Barton1979,Barton1986} have described patterns of clinal variation at loci linked to targets of selection.  
This provides some expectation for how selection shapes covariance in ancestry at a single locus over geographic space.
Our aim here is to extend this to include expectations for how selection shapes covariance in ancestry across the genome.}

@@

\paragraph{Establishment of the cline at the selected locus}
Deterministic theory predicts that after secondary contact,
the two alleles at the selected locus will form a stable cline \citep{Barton1979},
affecting neighboring loci as well.
First, we turn our attention to how the cline at the selected locus itself is formed.
Let $p(x,t)$ denote the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and that $p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
% (In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
Let $\sigma^2$ be the mean squared distance between parent and offspring 
(i.e., the variance of the distribution of dispersal distance).
Assuming that: (i) alleles locally assort into diploids randomly, 
(ii) habitat and dispersal are homogeneous, and 
(iii) $s$ and $\sigma$ are small, then the commonly-used equation that $p$ solves is
\begin{align} \label{eqn:cline_pde}
    \dot p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\dot p(t,x)$ is the time derivative and $\Delta p$ is the Laplacian,
as derived in \citet{Bazykin1969}; for discussion of diploidy see \citet{christiansen1995genotypic}.
Although there is no known exact analytic solution to this equation, 
the steady-state solution is
$\lim_{t \to \infty} p(x,t) = (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}. 
This relies on several approximations, but the general conclusions should be quite robust: 
the stable cline has width of order $\sigma/\sqrt{s}$, and decays exponentially. 
Note that rescaling space and time by $\sigma/s$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} results in a dimensionless equation,
implying that the cline establishes over a timescale of the order $1/s$. 
In other words, the cline is established through diffusion, despite being slowed down somewhat by selection against heterozygotes.
